---
title: "Diagnostic for ColocBoost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic for ColocBoost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates how to perform multi-trait colocalization analysis using diagnosic details in ColocBoost, 
including model parameters, model fitting, and model performance metrics. There is `diagnostic_details` in ColocBoost output when setting `output_level = 3`:


```{r setup}
library(colocboost)
# Run colocboost with diagnostic details
data(Ind_5traits)
res <- colocboost(X = Ind_5traits$X, Y = Ind_5traits$Y, output_level = 3)
names(res)
```

# 1. Diagnostic details of the model parameters

- **`cb_model_para`**: parameters used in fitting ColocBoost model.


```{r cb-model}
names(res$diagnostic_details$cb_model_para)
```


# 2. Diagnostic details of the model fitting

- **`cb_model`**: trait-specific proximity gradient boosting model, including proximity weight at each iteration, residual after gradient boosting, et al.
- **`weights_paths``**: individual trait-specific weights for each iteration.

```{r cb-model-para}
names(res$diagnostic_details$cb_model)
names(res$diagnostic_details$cb_model$ind_outcome_1)
```

