---
title: "Mixed Data-type and Disease Prioritized Colocalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed Data-type and Disease Prioritized Colocalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates how to perform multi-trait colocalization analysis using a mixed-type dataset, including both individual level data and summary statistics. 
ColocBoost provides a flexible framework to integrate data both at the individual level or at the summary statistic level, 
allowing to handle scenarios where the individual data is available for some traits (like xQTLs) and the summary data is available for other traits (disease/trait GWAS).


```{r setup}
library(colocboost)
```

# 1. Loading individual and summary statistics data

To get started, load both Ind_5traits and Sumstat_5traits datasets into your R session. Once loaded, create a mixed dataset as follows:

- For traits 1, 2, 3, 4: use individual-level gentype and phenotype data.
- For trait 5: use summary statistics data.
- Note that `LD` could be calculated from the `X` data in the `Ind_5traits` dataset, but it is not included in the `Sumstat_5traits` dataset.

### Causal variant structure
The dataset features two causal variants with indices 644 and 2289.

- Causal variant 644 is associated with traits 1, 2, 3, and 4.
- Causal variant 2289 is associated with traits 2, 3, and 5 (summary level data).

```{r load-mixed-data}
# Load example data
data(Ind_5traits)
data(Sumstat_5traits) 

# Create a mixed dataset
X <- Ind_5traits$X[1:4]
Y <- Ind_5traits$Y[1:4]
sumstat <- Sumstat_5traits$sumstat[5]
LD <- get_cormat(Ind_5traits$X[[1]])
```

For analyze the specific one type of data, you can refer to the following 
tutorals [Individual Level Data Colocalization](https://statfungen.github.io/colocboost/articles/Individual_Level_Colocalization.html) and 
[Summary Level Data Colocalization](https://statfungen.github.io/colocboost/articles/Summary_Level_Colocalization.html).


# 2. ColocBoost in disease-agnostic mode


The preferred format for colocalization analysis in ColocBoost using mixed-type dataset:

- **Individual level data**: `X` and `Y` are organized as lists, matched by trait index,
    - `(X[1], Y[1])` contains individuals for trait 1,
    - `(X[2], Y[2])` contains individuals for trait 2,
    - And so on for each trait under analysis.

- **Summary level data**:
    - `sumstat` is organized as a list of data.frames for all traits
    - `LD` is a matrix of linkage disequilibrium (LD) information for all variants across all traits.

This function requires specifying genotypes `X` and phenotypes `Y` from the individual-level dataset and summary statistics `sumstat` and LD matrix `LD` from summary dataset:


```{r mixd-basic}
# Run colocboost
res <- colocboost(X = X, Y = Y, sumstat = sumstat, LD = LD)

# Identified CoS
res$cos_details$cos$cos_index

# Plotting the results
colocboost_plot(res)
```

### Results Interpretation

For comprehensive tutorials on result interpretation and advanced visualization techniques, please visit our documentation portal 
at [Visualization of ColocBoost Results](https://statfungen.github.io/colocboost/articles/Visualization_ColocBoost_Output.html) and
[Interpret ColocBoost Output](https://statfungen.github.io/colocboost/articles/Interpret_ColocBoost_Output.html).



# 3. ColocBoost in disease-prioritized mode

Disease-prioritized mode of ColocBoost can 

When integrating with GWAS data, to mitigate the risk of biasing early updates toward stronger xQTL signals in LD proximity—and 
should in fact colocalize with—weaker signals from the disease trait of interest, 
we also implement a disease-prioritized mode of ColocBoost as a soft prioritization strategy to colocalize variants with putative causal effects on the focal trait.

To run the disease-prioritized mode, you need to specify the `focal_outcome_idx` argument in the `colocboost()` function.

- `focal_outcome_idx` indicates the index of the focal trait in all traits of interest.
- Note that the `focal_outcome_idx` are counted from *individual level data* to *summmary statistics*.
Therefore, in this example, `focal_outcome_idx = 5` is used to indicate the index of the focal trait from (4 individual level traits) and (1 summary statistcs).


```{r disease-basic}
# Run colocboost
res <- colocboost(X = X, Y = Y, 
                  sumstat = sumstat, LD = LD, 
                  focal_outcome_idx = 5)

# Plotting the focal only results colocalization results
colocboost_plot(res, plot_focal_only = TRUE)
```

Unlike the other existing methods, the disease-prioritized mode of ColocBoost only used a soft prioritization strategy.
Therefore, it does not only identify the colocalization of the focal trait,  but also the colocalization across the other traits without the focal trait. 
To extract all CoS and visualization of all colocalization results, you can use the following code:

```{r all-basic}
# Identified CoS
res$cos_details$cos$cos_index

# Plotting all results
colocboost_plot(res)
```

### Results Interpretation

For comprehensive tutorials on result interpretation and advanced visualization techniques, please visit our documentation portal 
at [Visualization of ColocBoost Results](https://statfungen.github.io/colocboost/articles/Visualization_ColocBoost_Output.html) and
[Interpret ColocBoost Output](https://statfungen.github.io/colocboost/articles/Interpret_ColocBoost_Output.html).