---
title: "Input Data Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input Data Format and Example Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(colocboost)
```

## Input Data Format

This vignette documents the required input data formats and provides examples of data included in the package.

### Individual Level Data

For analyses using individual-level data, the package requires matched X and Y data. Below is the format and an example from the package:

```{r individual-level-example}
# Load example individual-level data
data(Ind_5traits)

# Display the structure
str(Ind_5traits)
```

#### Format Requirements

- Data should be in a data frame or matrix
- Each row represents an individual
- Columns must include matched genotype (X) and phenotype (Y) data
- Missing values should be coded as NA

### Summary Statistics

For analyses using summary statistics, the package requires a data frame with matched linkage disequilibrium (LD) information:

```{r summary-stats-example}
# Load example summary statistics data
data(Sumstat_5traits)

# Display the structure
str(Sumstat_5traits)
```

#### Format Requirements

- Data should be in a data frame
- Each row represents a variant
- Must include effect size, standard error, and sample size information
- LD matrix must be provided separately or calculated from the data
